\documentclass{article}

\usepackage{biblatex}

\addbibresource{ref.bib}

\title{ELE 548: Project Proposal}

\author{Calvin Higgins}

\begin{document}

\maketitle

% Submit by Nov 20th. Progress reports should include a revised version of the proposal (if necessary).
% Primarily I am looking for a one to two pages document describing accomplishments so far and a weekly
% plan of work for the remainder of the semester. Concentrate on describing sub-tasks completed, rather than
% the tasks started.
% Your report should include the following:
% + What parts of design specification are complete
% + Design implementations status. For example, if you are modifying a simulator, what mods are
% done
% + Outline of planned experiments. This can evolve and change, but I want you to have thought of
% an initial plan.
% + Finally, whatever questions/comments I marked on the proposal (or we have discussed during our
% project meetings) must be addressed. Include a response to the questions

\section{Completed Tasks}

\subsection{Learn Reinforcement Learning}

As I have no prior reinforcement learning experience, I completed a reinforcement learning tutorial from the \texttt{pytorch} documentation. I trained a deep 
Q-network (DQN) to balance a pole on a cart by moving left or right. 

\subsection{Install CompilerGym}

I installed CompilerGym~\cite{cummins2022}, the standard framework for reinforcement learning in compilers. This was surprisingly difficult: CompilerGym is not 
actively maintained, many of its requirements are deprecated, and compilers are generally challenging to install correctly. I got CompilerGym working with 
\texttt{docker} using GPU passthrough and the \texttt{uv} package manager. 

\subsection{Implement DQN}

I tried to keep my initial implementation simple and fast. I implemented an epsilon-greedy DQN and targeted IR instruction count reduction. I used Autophase 
features~\cite{hajali2020} extracted from the IR code as the program representation. Since Autophase features are integer values with high variance, I 
standardized them with Welford's online algorithm. 

\subsection{Run Initial Experiments}

I verified my implementation by optimizing IR instruction counts on several benchmark suites. Again, to keep the initial implementation simple and fast, I did 
not use a held-out test set. After (painful amounts of) tuning, I was able to successfully overfit a single benchmark (quicksort), the CHStone 
suite~\cite{hara2008} (12 benchmarks) and the cBench suite (23 benchmarks)~\cite{fursin2014}, outperforming \texttt{-Oz} on average. I was not able to 
outperform \texttt{-Oz} on the MiBench suite~\cite{guthaus2001} (40 benchmarks). Initially, I thought that my model was struggling to learn as the number of 
programs increased, however, it turns out that it's just very hard to beat \texttt{-Oz} on MiBench~\cite{chaoyi2025}. After realizing this, I tried the 
Tensorflow suite~\cite{abadi2016} (1985 benchmarks), and beat \texttt{-Oz} on average. Notably, my training run crashed because the size of an optimized program
became zero, indicating either (1) a compiler bug or (2) a Tensorflow bug. Unfortunately, I did not record enough data to determine which actually happened.

\section{Remaining Tasks}

\subsection{Monitoring Infrastructure}

I implemented some infrastructure for monitoring relevant metrics such as gradient norms and losses. However, before moving on, I want to add monitoring for 
the reward, action, Q-value, and IR instruction count distributions to inform later design decisions. I estimate this will take around one day of work.

\subsection{Feature Representation}

I believe the feature representation can be greatly improved. I have several ideas I want to try including normalizing inputs and rewards to the baseline 
state, pretrain an autoencoder on Autophase features, and try an Transformer encoder with inst2vec embeddings~\cite{bennun2018}. I estimate this will take 
around one week of work.

\subsection{Evaluation}

Currently, I am not evaluating on a held-out test set. This is bad for comparison but fine for experimentation since phase-ordering is very hard to overfit. I 
need to add evaluation on a held-out test set. I estimate this will take around one day of work. 

\subsection{EfficientZero}

I need to replace the DQN with EfficientZero~\cite{ye2021}. I estimate this will take around one week of work.

\printbibliography

\end{document}